---
title: "Map the Life Expectancy in United States with data from Wikipedia with R"
author: "Andrei Macsin"
date: "September 2, 2016"
output: 
  html_document: 
    highlight: pygments
    theme: cerulean
---

```{r, echo=FALSE}
setwd("d:/tmp/Dropbox/Edu/Kaggle/various")
rm(list=ls(all=TRUE))

library(knitr)
opts_chunk$set(echo = TRUE, cache = T, cache.path = "cache/", fig.path = "figure/", warning = FALSE)
#http://yihui.name/knitr/options/
```

This is code from [DataViz](http://www.datavizualization.com/blog/map-the-life-expectancy-in-united-states-with-data-from-wikipedia).


```{r, echo = F}
library(rvest)
library(ggplot2)
library(dplyr)
library(scales)
```

# Load the data

Note a very intersting sign `%>%`, as describedin [this stack overflow post](http://stackoverflow.com/questions/24536154/what-does-mean-in-r) this is de-facto unix pipe. In other words it is passing value from left to right
```{r}
library(magrittr)
le = read_html("https://en.wikipedia.org/wiki/List_of_U.S._states_by_life_expectancy")
le = le %>% html_nodes("table") %>% .[[2]]%>% html_table(fill=T)

```


# Clean and transform the data
```{r}
# select only columns with data
le = le[c(1:8)]
# get the names from 3rd row and add to columns
names(le) = le[3,]
# delete rows and columns which I am not interested
le = le[-c(1:3), ]
le = le[, -c(5:7)]
# rename the names of 4th and 5th column
names(le)[c(4,5)] = c("le_black", "le_white")
# make variables as numeric
le = le %>%
mutate(
le_black = as.numeric(le_black),
le_white = as.numeric(le_white))
```

Since there are some differences in life expectancy between White and African-American, I will calculate the differences and will map it.
```{r}
le = le %>% mutate(le_diff = (le_white - le_black))
```

# Load the map skin

```{r}
states = map_data("state")
# create a new variable name for state
le$region = tolower(le$State)
# merge the datasets
states = merge(states, le, by="region", all.x=T)
```

# Plot the data

```{r}
ggplot(states, aes(x = long, y = lat, group = group, fill = le_black)) +
geom_polygon(color = "white") +
scale_fill_gradient(name = "Years", low = "#ffe8ee", high = "#c81f49", guide = "colorbar", na.value="#eeeeee", breaks = pretty_breaks(n = 5)) +
labs(title="Life expectancy in African American") +
coord_map()
```

