---
title: "Using segmented regression to analyse world record running times"
author: "Andrie de Vries"
date: "December 31, 2015"
output: 
  html_document: 
    highlight: pygments
    theme: cerulean
---

```{r, echo=FALSE}
setwd("d:/tmp/Dropbox/Edu/Kaggle/various")
rm(list=ls(all=TRUE))

library(knitr)
opts_chunk$set(echo = TRUE, cache = T, cache.path = "cache/", fig.path = "figure/", warning = FALSE)
#http://yihui.name/knitr/options/
```

This is code from [Rev article](http://blog.revolutionanalytics.com/2015/12/using-segmented-regression-to-analyse-world-record-running-times.html). Git snippet can [be found here](https://gist.github.com/andrie/832f9ab09e4a3d23d0d5).


# Clean and transform the data

```{r, echo = F}
dat <- structure(list(Distance = c(0.1, 0.2, 0.4, 0.8, 1, 1.5, 1.6, 
2, 3, 5, 10, 15, 20, 21.1, 25, 30, 42.2, 100, 50, 100, 160.9), 
    Time = c("00:00:10", "00:00:19", "00:00:43", "00:01:41", 
    "00:02:12", "00:03:26", "00:03:43", "00:04:45", "00:07:21", 
    "00:12:37", "00:26:18", "00:41:13", "00:56:26", "00:58:23", 
    "01:12:25", "01:26:47", "02:02:57", "06:13:33", "02:48:06", 
    "06:10:10", "11:28:03"), Athlete = c("Usain Bolt", "Usain Bolt", 
    "Michael Johnson", "David Rudisha", "Noah Ngeny", "Hicham El Guerrouj", 
    "Hicham El Guerrouj", "Hicham El Guerrouj", "Daniel Komen", 
    "Kenenisa Bekele", "Kenenisa Bekele", "Leonard Patrick Komon", 
    "Haile Gebrselassie", "Zersenay Tadese", "Moses Mosop", "Moses Mosop", 
    "Dennis Kipruto Kimetto", "Takahiro Sunada", " Jeff Norman (GBR)", 
    " Donald Ritchie (GBR)", " Oleg Kharitonov (RUS)"), Date = c("16-Aug-09", 
    "20-Aug-09", "26-Aug-99", "09-Aug-12", "05-Sep-99", "14-Jul-98", 
    "07-Jul-99", "07-Sep-99", "01-Sep-96", "31-May-04", "26-Aug-05", 
    "26-Sep-10", "27-Jun-07", "21-Mar-10", "03-Jun-11", "03-Jun-11", 
    "28-Sep-14", "21-Jun-98", "07-Jun-80", "28-Oct-78", "20-Oct-02"
    ), Record = c("World record", "World record", "World record", 
    "World record", "World record", "World record", "World record", 
    "World record", "World record", "World record", "World record", 
    "World record", "World record", "World record", "World record", 
    "World record", "World record", "World record", "World best", 
    "World best", "World best"), Time.in.hours = c(0.0027, 0.0053, 
    0.012, 0.028, 0.0367, 0.0572, 0.062, 0.0791, 0.1224, 0.2104, 
    0.4382, 0.6869, 0.9406, 0.9731, 1.2071, 1.4465, 2.0492, 6.2258, 
    2.8017, 6.1694, 11.4675), Standard.event = c(" Yes ", " Yes ", 
    " Yes ", " Yes ", " Yes ", " No ", " Yes ", " No ", " Yes ", 
    " Yes ", " Yes ", " No ", " No ", " Yes ", " No ", " No ", 
    " Yes ", " No ", " No ", " No ", " No ")), .Names = c("Distance", 
"Time", "Athlete", "Date", "Record", "Time.in.hours", "Standard.event"
), class = "data.frame", row.names = c(NA, -21L))
```



```{r}

library(ggplot2)
#dat <- read.csv("world records.csv", stringsAsFactors = FALSE)

track <- within(dat, {
  Time <- as.numeric(Time.in.hours)
  Date <- as.Date(Date, format = "%d-%b-%y")
  Speed <- Distance / Time
  logDistance <- log10(Distance)
  Standard.Event <- Standard.event == " Yes "
  Time.in.hours <- NULL
  Standard.event <- NULL
})


```

# Understanding data


```{r}
ggplot(aes(y = Speed, x = Distance), data = track) +
  geom_point() +
  theme_bw(16) +
  ggtitle("World record speed for mens running events")
```

You can immediately see that the speed declines very rapidly from the sprint events. Perhaps it would be better to plot this using a logarithmic x-scale, adding some labels at the same time. 
I also added some colour for what I call standard events - where "standard" is the type of distance you would see regularly at a [world championships](http://www.iaaf.org/competitions/iaaf-world-championships) or [olympic games](http://www.iaaf.org/competitions/iaaf-world-championships). Thus the mile is "standard", but the 2,000m race is not.


```{r}

labelx <- c(0.1, 0.2, 0.8, 1.6, 3, 10, 21.1, 42.2)
labely <- track[track$Distance %in% labelx, "Speed"]
labeltext <- c("100m", "200m", "800m", "Mile", "3,000m", "10,000m", "Half-marathon", "Marathon")
labels1 <- data.frame(Distance = labelx, Speed = labely, Text = labeltext)

labelx <- c(50, 100, 160.9)
labely <- track[track$Distance %in% labelx & track$Record == "World best", "Speed"]
labeltext <- c("50km", "100km", "100mile")
labels2 <- data.frame(Distance = labelx, Speed = labely, Text = labeltext)


ggplot(aes(y = Speed, x = Distance), data = track) +
  # geom_path(data = data.frame(Distance = newx, Speed = newy), colour = "grey 20") +
  geom_point(aes(colour = Standard.Event), size = 3) +
  geom_text(data = labels1, aes(label = Text), hjust = -0.2, vjust = -0.2) +
  geom_text(data = labels2, aes(label = Text), hjust = 1.2, vjust = -0) +
  scale_x_log10() +
  #theme_bw(16) +
  ggtitle("World record speed for mens running events")


```

Now our data points are in somewhat more of a straight line, meaning we could consider fitting a linear regression. However, it seems that there might be two kinks in the line:

* Btween the 800m distance and the mile. It seems that the sprinting distances (and the 800m is sometimes called a long sprint) has different dynamics from the events up to the marathon.
* another kink for the ultra-marathon distances. The standard marathon is 42.2km, and distances longer than this are called ultramarathons.

The *segmented()* function allows you to modify a fitted object of class lm or glm, specifying which of the independent variables should have segments (kinks).  In my case, I fitted a linear model with a single variable (log of distance), and allowed segmented() to find a single kink point.

```{r}

library(segmented)

# Helper function to compute the inverse of log10(x)
exp10 <- function(x)10^x

# Limit model data to distance from 200m to marathon
modeldata  <- track[track$Distance >= 0.2 & track$Distance <= 42.2, ]

# Fit linear model
lfit <- lm(Speed ~ logDistance, data = modeldata)

# Fit segmented model
sfit <- segmented(lfit, seg.Z = ~ logDistance)
exp10(sfit$psi)
summary(sfit)

bpoints <- 10^(sfit$psi)[, 2]
newx <- c(0.2, bpoints, 50)
newy <- predict(sfit, data.frame(logDistance = log10(newx)))
```


Lets visualise this

```{r}
labelx <- c(0.1, 0.2, 0.8, 1.6, 3, 10, 21.1, 42.2)
labely <- track[track$Distance %in% labelx, "Speed"]
labeltext <- c("100m", "200m", "800m", "Mile", "3,000m", "10,000m", "Half-marathon", "Marathon")
labels1 <- data.frame(Distance = labelx, Speed = labely, Text = labeltext)

labelx <- c(50, 100, 160.9)
labely <- track[track$Distance %in% labelx & track$Record == "World best", "Speed"]
labeltext <- c("50km", "100km", "100mile")
labels2 <- data.frame(Distance = labelx, Speed = labely, Text = labeltext)


ggplot(aes(y = Speed, x = Distance), data = track) +
  geom_path(data = data.frame(Distance = newx, Speed = newy), colour = "grey 50", size = 2) +
  geom_point(aes(colour = Standard.Event), size = 3) +
  geom_text(data = labels1, aes(label = Text), hjust = -0.2, vjust = -0.2) +
  geom_text(data = labels2, aes(label = Text), hjust = 1.2, vjust = -0) +
  scale_x_log10() +
  theme_bw(16) +
  ggtitle("World record speed for mens running events")

```

# Conclusion

I conlude:

* There seems to be a different physiological process for the sprint events and the middle distance events. The segmented regression finds this kink point between the 800m event and the 1,000m event
* The ultramarathon distances have a completely different dynamic. However, it's not clear to me whether this is due to inherent physiological constraints, or vastly reduced competition in these "non-standard" events.
* The 50km world record seems too "slow". Perhaps the competition for this event is less intense than for the marathon?

